Conservator's Corner

### 2023-05-09 pnsac/newsletter/article_MemberInterview.tex

\title{Our Members}
% \author{Interview with Garry Dupont}

@@

@@

Readers of the Chronicle will know that this item usually refers to one of our
members. This edition, however, salutes all of our members who worked as
volunteers on the restoration of the North Star engines namely:

\begin{itemize}
  \item Charles Baril 
  \item John Corby (deceased)
  \item Ted Devey (deceased)
  \item Ed Hogan
  \item Peter Houston
  \item Ron Lemieux
  \item Richard Lodge
  \item Wib Neal
  \item Jim Riddoch
  \item Stan Rideout
  \item Ted Slack
  \item John Tasseron
  \item Bill Tate
  \item Tim Timmins
  \item Giuseppe Zanetti
\end{itemize}

The following members have provided comments on their personal experiences.

\textbf{Charles Baril}

Working on all 4 Merlin engines has been an extremely rewarding experience that
has taught me a lot about older technology, an appreciation for history
involved (and those in the engine crew with that knowledge) and learning new
skills and techniques. Although most of the parts were auxiliary systems,
seeing them as part of the full engine gives a sense of accomplishment. It is
very satisfying to see a part going from the dirty, oil covered, rusty part to
newly refurbished one that looked like it was overhauled in an RCAF maintenance
depot. It has been an honour to be part of this project and restoration of
these pieces of history.

In particular I remember working on the lower cowling flaps.  Trying to undo
the inner skin was difficult due to longer rivets through the mounts for the
Flaps. It took several tries and a few extra holes, but it was done. Fixing
those holes often meant using original methods which was revealing and a
commitment to historical methods.

I also recall the work on the filter ducts.  The filter duct is the last stage
of the air intake before the supercharger. During its cleaning, there was a
starling nest on top of one filter. This caused a lot of corrosion damage that
needed to be removed and treated. Yet, what got me was that starlings could get
that far into the engine.

\textbf{Richard Lodge}

As probably the only volunteer who ever worked in the Aero Engine Division of
Rolls-Royce in Derby, England, I was greatly interested in becoming involved in
Project North Star, even though, as an accountant, I had never worked on the
design or building of an aeroengine.  By the time I started work at Rolls-Royce
in 1963, new build of Merlin engines had ceased, and only spare parts were
being manufactured.

I have several strong memories of working on the Merlin engine.  In general,  I
particularly remember John Tasseron working endlessly on freeing seized piston
rings.  I learned much from John about patiently working at something which
appeared to be impossible! Later, when working with  R\'{e}j Demers and Garry
Dupont, I started to learn how to rivet and how to lock wire. The most
important lesson for me has been, photograph, photograph, photograph, at each
stage of dismantling, cleaning and reinstalling an engine part. When sitting in
an overcrowded office at Rolls-Royce, it never occurred to me that at some time
in the future I would actually be working on a Merlin engine or working my way
through an original parts manual to work out how to reassemble an item.  Life
is certainly full of interesting surprises.

\textbf{John Tasseron}
 
The list below is intended to give a brief overview of the various highlights
that arose during my participation in the restoration of the second, third and
fourth Merlin engines:

\begin{enumerate}
  \item It's not every day one gets the opportunity to see up close what makes
the Rolls-Royce 620 Series Merlin engine tick!
  \item It's a real challenge to safely lower each "power egg" down off the
wing, place it in "the big blue stand" and move it in to the workshop.
  \item Then starts the process of carefully separating about 11,000 individual
parts before cleaning, inspecting and re-conditioning them as necessary. Do not
lose any and make sure everything is photographed for museum records!
  \item When things unexpectedly go wrong, the real expertise kicks in to
perhaps develop a repair scheme or even manufacture a brand-new replacement
part. The objective is to bring everything back together again and end up with
a "like new" engine, ready for preservation.
  \item It all takes team effort based on effective communication and
co-operation. There are no dumb questions and the answers are all aimed at the
challenge to create an "airworthy" product that will never fly.
  \item Finally, its important and fun to share the project with a curious
public, so be ready to answer any questions that might be asked about the
aircraft and its engines!
\end{enumerate}

>>>>

\title{Our Members}
\author{Richard Lodge.}

@@

Last New Year's Eve, December 2021, I became the third past president of PNSAC, after 12 years in the position and following several years at treasurer. I was delighted when Chris McGuffin agreed take on the role of president following the unanimous support of the board of directors.

Before starting to write this article, I made a list of some of the more important and memorable events of my presidency. This article is my recollection of the things that have stuck in my mind over the years. During my term in office, PNSAC changed from being an experimental volunteer organization which still had to prove itself to the Canadian Aviation and Space Museum (CASM), to being part of the CASM family where PNSAC members are well respected and trusted by the museum staff and management.

Several events stand out when I think about my years as president. The first was on Canada Day 2017 shortly after R\'{e}j Demers had arrived at CASM to take over the position of Special Project Manager. The ex RCAF Hercules C130 had arrived in April 2016 and was parked on the apron outside the Storage Hangar.  A large crowd gathered around the aircraft when the day was at its it hottest.  Standing in the middle of the cargo ramp was R\'{e}j where for half an hour or so he entertained the crowd.  The event was unplanned, and his comments were unscripted. The crowd just stood there enjoying the informality of listening to what he was saying and answering questions shouted to him.

The second event was in December 2019 when Engine \#4 was moved from the Engine Shop to be located in front of the North Star in the Storage Hangar. It was a cold early winter day with snow flurries. On this occasion the engine could only be moved on its stand which had four wheels and no steering. It was towed by R\'{e}j driving CASM's forklift from the staff parking lot along the side of the building and into the Storage Hanger. It was a very slow process taking something like half an hour. By the time we got the engine safely to bed the three engine shop volunteers (me, Garry Dupont, and Charles Baril) as well as other volunteers who were walking behind the engine were really not enjoying the beginning of a Canadian winter. R\'{e}j who was sitting in the heated cab of the forklift, with a nice smile on his face, was actually quite warm!

\begin{figure}[H]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.75]{engin_4.png}
   %caption of the figure 
   \caption*{\small \em From left to right: Garry Dupont, R\'{e}j Demers (in the cab), and Chris McGuffin.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:img1}
\end{figure}

The next event was when I was MC at the showing of the film  made in 2014 when the  Lancaster bomber, part of the collection of the Canadian Warplane Heritage Museum, flew to England to join the only other airworthy Lancaster in the world for a summer tour of the airshows in Britain. The epic flight was filmed throughout its three-day journey to England, during its time in England and on its return.  Bill Tate, one of our very active members, arranged for the film to be shown in the auditorium of the CASM to about 250 members of the general public.  The film itself was very interesting, but what made it even more interesting was that the flight crew agreed to come to Ottawa and talk to the audience after the showing of the film.  Hearing the flight crew talk of the preparations and difficulties of the tour and knowing how successful it had been was particularly interesting to me.  I had managed to attend the first airshow in England when the two Lancasters, a Hurricane and two Spitfires had flown together off the south coast of England, near where they would regularly pass on bombing missions during World War 2. The sound of 11 Merlin engines flying together reminded me of my boyhood in England when I would hear the bombers leaving their bases in Yorkshire to bomb Germany – quite an emotional sound all these years later.

There are always difficulties and problems to deal with when one is the head of any organization, be it in a paid position or as a volunteer.  PNSAC has experienced a  number of challenges during it existence.  We have had two major closures and a lock down and closure caused by the Covid 19 pandemic.  The Covid 19 closure needs no explaining but the previous two were necessitated by CASM staging major events in 2013 and 2016 around the Storage Hangar.  While the closures were disruptive to the Project North Star operations, the presentation of the Star Wars and Star Trek exhibits allowed CASM to reach a broader audience and raise awareness of CASM. The exhibits made it impossible for the restoration team to work on the North Star aircraft for many weeks in both years.  The challenge was to try to keep the restoration volunteers together when some volunteers were finding other ways to occupy their time.  Many meetings and discussions took place, and another side of my role was to help the restoration volunteers understand that these closures were only temporary and that as soon as the spectacle events were finished, restoration work would start on the aircraft again.  Fortunately, we did not lose many restoration volunteers although our actual non-volunteer membership dropped considerably.

During the summer of 2018 it was decided that the vertical fin of the North Star should be removed.  This involved weeks of preparatory work inside the aircraft, loosening the many nuts and bolts securing the fin to the aircraft.  On the appointed day a large mobile crane arrived to remove the vertical fin. The rental cost was financed by PNSAC donations received from members.  It was a hot day, and the crane was duly stationed beside the aircraft.  However, several of the securing bolts had been left in place and needed to be removed once the crane had been attached to the fin.  Several volunteers assisted in the attaching of the crane hook to the fin, but one volunteer, Charles Baril, offered to crawl inside the plane and remove the last remaining bolts.  The inside of the plane was like an oven with no ventilation and no proper lighting. The only way Charles could communicate with the crew outside was by shouting.  After about an hour of very difficult work, Charles shouted that all the nuts and bolts had been undone and the actual lifting of the fin could proceed.  After the fin had been lifted by the crane, it was lowered into a specially prepared cradle. Charles emerged from the aircraft to the applause of all people outside, looking as if he was about to melt and thoroughly tired out by his exertions.

\begin{figure}[H]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{nstar_tail_fin.png}
   %caption of the figure 
   \caption*{\small \em The North Star with its tail fin removed.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:img1}
\end{figure}

 As president, I was involved in several other memorable events organized by Bill Tate.  Bill has an amazing ability to organize events and had a substantial list of contacts.  I was primarily involved as the MC of the trips while in the buses, also providing a backup for Bill should he encounter any problems.  Twice we visited RCAF Trenton and once we went to Montréal and Mirabel to see the Canadair/Bombardier production line in action. The most memorable was the weekend bus trip to Hamilton to attend the airshow organized by the Canadian Warplane Heritage Museum on the Saturday and on the Sunday, a visit to the Air Canada control centre in Toronto. On the Friday evening, Bill organized a visit to the vineyards in Niagara-on-the-Lake. Sunday morning saw us all back in the bus heading for downtown Toronto, where we breakfasted in the brewery district before our Air Canada visit.   Many members will have great memories of these times, which all took place prior to the closures I referred to earlier in this article.

\end{multicols}

\begin{figure}[H]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.4]{july2012.png}
   %caption of the figure 
   \caption*{\small \em Hamilton trip.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:img1}
\end{figure}

\begin{multicols}{2}

One of the best aspects of assuming responsibility as a volunteer is that the more one puts into it, the more pleasure and interest one gets from the work.  As president, I regularly met all the senior management of CASM and many other people outside CASM.  Often, I was involved in discussions about the future direction of PNSAC, which of course involved discussing future and sometimes confidential CASM plans and operations.

A very small incident, probably four or five years ago, really demonstrated to me how as a volunteer association we had "arrived".  My predecessor presidents, Bob Holmgren and Tim Timmins had to overcome the skepticism of CASM management as to whether it would be possible for us to sustain a continuing volunteer effort.  By the time I became president in 2010, CASM management had accepted that our members would work to a very high standard of restoration work and would continue to put in several thousand hours of painstaking and sometimes difficult work every year.  The incident I am remembering best is a particular day when I went upstairs to the office level  at CASM and saw a group of CASM staff talking in the corridor.  I stopped and said hello to them and then started to excuse myself by saying that they were having a discussion and I would get out of the way.  Chris Kitzan, the CASM Director General, looked at me and said "There was no need to go away Richard, you are part of the family".  This for me was a watershed moment and one I shall always appreciate.

Looking towards the future, as the pandemic begins to be contained, we must get the restoration operation going again and try to recruit more members and restoration volunteers.  I feel very confident that Chris McGuffin and the other directors of PNSAC will be able to do this and continue to help CASM in volunteer work in addition to North Star restoration.  I look forward to continuing volunteering, but definitely not climbing around the top of the aircraft in a full safety harness!

@@

### 2023-05-09 pnsac/newsletter/article_ProgReport.tex

When touring the shops in the Museum's Conservation wing, the engine shop is
reserved for what is referred to as, the "Crown Jewel". Guests are lead through
the corridor and the lights are flicked on to reveal a shining piece of
aeronautical engineering and craftsmanship. The Merlin 622 engine, one of four
such power plants, sits resplendent. Gasps and expressions of delight from
guests are guaranteed to insight a prideful glance upon this piece of work by
volunteers and staff alike. Appearing as if the engine had just been delivered
from Bristol Aero Ltd. is an illusion. What is this strange magic? How can such
beauty manifested though steel and aluminum even exist? As any good sorcerer
will tell, illusion is lost upon the dissolution of mystery. Yet this piece of
work, a restoration of great effort, time and detail, cannot shed the cloak of
timelessness so easily. The result of thousands of hours culminate through
unrelenting intrigue, garnered by all who cast eyes upon the Merlin. These
words simply scatter as embellishments to the enduring design and purpose,
which cannot be hidden in all evidence of its greatness.

Referencing an Ottawa Citizen article dated Saturday Feb. 1st, 2003. We read of
an effort to restore the Canadair North Star to a version of her "former self"
as PNSAC founder Robert Holmgren put it. A sad, sorry state at the time.
Destitute to serve as a bird house to the starlings. No mention of any spells
enacted to summon the aviation gods or mojo hand placed upon the yoke to stir
the spirits which have tilled the air.  Yet today, she's a Hangar Queen,
patiently awaiting final installation of engine \#4. Is this sorcery? No,
simply the fourth iteration of the Project's labour of love; Rolls-Royce Merlin
622, Serial \#309573. The task of accomplishing the finale to a four part, long
haul repeat restoration of a power plant began in 2014. Looking back at power
plant restoration cycles, we see the evolution of the Project North Star
volunteer organization since the early 2000's. The forming of Engine Shop crews
and the establishing of Crew Chiefs. The coordination of daily grunt work of
condition recording, treatment and reassembly of components. Today, members of
the project continue to develop ideas, adapting to new technology, finding
different approaches and maintaining a solid work ethic. This spirit remains
unchanged throughout difficult times and challenges we have encountered.

The Tech. log states: Engine installed on North Star 17515 on August 23rd, 1965
at airframe Total Time of 19845.45 Hrs. with 1121 hours remaining in \#4
position (4th to the right). The engine had previously flown on 17514 until
April 30th, 1965 when a broken intake valve \@ \#1 cylinder, "B" bank
necessitated a repair. Flipping through pages, various signatures, authorities,
stamps and markings show evidence of a storied service life going back to 1948.
Engines are shared among power plant "skeletons" which in turn, are shared
across the fleet. This was one of the major design features of the "Power Egg"
as the promotional material of the day had put it, interchangeability and ease
of maintenance. A glance at the log tells me this engine serial number served
on North Stars 506, 510, 508 and 502 to name a few. The last remaining example
of its type, North Star 17515 holds pieces of the past, belonging to aircraft
that have long been gone. Like crew members who have also shared their time
amongst a fleet that traveled all over the globe, with change-overs,
turn-arounds and plenty of distance in-between.

One of the big pay-offs for members of Project North Star is the opportunity to
display the work carried out by volunteers. Many occasions come to mind, such
as Canada Day, Doors Open Ottawa and family visits to the Museum. One
opportunity, of which I will always be grateful to have participated in, was
such a moment. Myself and a few volunteers met a retired employee of the
aforementioned Bristol Aero Industries Ltd. who worked at the overhaul shops
when this engine was undergoing acceptance runs in the early to mid-sixties.
Quickly, job experience and familiarity with the object was brought forward. A
quick look under the cowling to reveal preserved markings and decals, evidence
of original repair shop work was pointed out. "I probably worked on that job"
he said. I understood at this point, how someone's work, preserved in such a
way, could be of intangible value. The history coming around full-circle before
my eyes, as the gentleman pointed out the ROLLS-ROYCE cast upon the rocker
covers, saying "it was one of my first jobs, painting on those red letters".

These records, object qualities, historic log entries and our treatment reports
all add up. We string together these bits of information to form longer chains
that communicate the history of an artifact. As focal points shift towards
other areas of interest, or priorities, work carried out in the past helps
maintain a continued relevance for others to pick-up where we left off. With
such a long term project, setting the pace is just as important as taking a
break now and then. Knowing that we had an impact and remembering the
experiences we've shared can carry us through new challenges. It's a pleasure
for me to observe a new volunteer work towards acquiring the skills that may be
passed on to others, like an old wizard speaking to his young apprentice.

I encourage Project North Star members to communicate their interest in the
project to the public. Generate their own records that may contribute to the
history of the machine and the lives it affected. This newsletter, social media
and other forms of reaching out to the public are excellent ways to remember
one of Canada's great contributions to air transport. With the tools at our
disposition, I would not hesitate to believe we can create many new and
engaging experiences for the public to learn about the role this aircraft
undertook while in service. Now, in continuation with a sense of national
pride, the North Star remains an important platform for interpretation of the
work carried out by dedicated volunteers supported by a museum that understands
the value of such an artifact.

>>>>

National Volunteer Week was held recently. It was a timely call to action. Let's focus our collective strengths and resources to prepare for Project North Star RESTART 2022.

It has been roughly 20 years now, since the beginning of the North Star restoration project. Over the past two years, we've had to endure the effects of the CoVID-19 global pandemic. This spring, we have an opportunity to make an attempt to (gradually) return to normal operations. Ingenium is currently initiating actions that will permit the return of Project North Star volunteers. This will require a concerted effort on behalf of all stakeholders.

My intent is to have crews receive training, gradually integrating in numbers, until we could finally support full daily compliments based on operational capacities. Now, with the demands imposed by Ingenium's recent acquisition approval of an RCAF CC-115 Buffalo, we must act upon this news in conjunction with a collection re-org. This signifies an earlier than expected requirement for Project North Star volunteer support, due to the increased scope of work.

CASM Conservation is preparing a work schedule reliant upon this volunteer support, in order to mobilise the North Star from the Reserve Hangar.  Moving forward, through consultations with all parties, obtaining guidelines and limitations, as well as setting goals and expectations so that we may provide an integrated work plan for 2022-2023.

High on our list of immediate concern is rendering 17515 mobile so that we may conduct a massive hangar shuffle. Collection objectives are driving CASM Conservation Services (and Project North Star) into a 24 month action plan which will eventually see us dispose of some artifacts and acquire one big yellow beast (115452). Thankfully, I have been able to secure accommodation within the storage hangar for the North Star between brief stints outdoors during fair weather movements, in amongst the storage hangar and the main museum.

Project North Star has an important role in reinvigorating the museum ecosystem, providing a necessary boost of youthful vigour and effervescence. You may be asking yourself if during this prolonged absence, a time machine or temporal portal was made available for us to return to a simpler time, a golden age of aviation perhaps? Disappointingly, such a device has not yet surfaced beyond the top secret labs of the NRC. We must rely upon more tangible resources, such as students, cadets and young urban professionals. Diversification and integration of alternative human resources are being tapped, like springtime sugar maples.

It's been a long winter, or two. Attempts to grasp fleeting moments of human connection leave us questioning what sort of catharsis may be obtained from a grocery store checkout aisle. I myself, have taken steps towards awkward avenues of discussion by making cold-calls to random organizations which may prove to be valuable partners in the advancement of the project. Recognizing the North Star as an important platform for discussion and broadening the scope of Canadian heritage interpretation.

With this recent streak of renewal, and a strong and healthy relationship, Project North Star and in turn, Ingenium may remain viable organizations as we continue our recovery. I look forward to hearing back from everyone, so that we may set a date for Project North Star RESTART 2022.
